---
title: "Sb to SB differences heatamp"
author: "Carlos Martinez Ruiz"
date: "18 January 2019"
output: pdf_document
---

```{r setup, include = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE)#, fig.path = 'results/')
```
```{r libraries, echo = FALSE, message = FALSE, warning = FALSE}
#Load all libraries
library(tximport)
library(readr)
library(DESeq2)
library(GenomicRanges)
library(GenomicFeatures)
library(ggplot2)
library(plyr)
library(lmerTest)
library(quantreg)
library(eulerr)
version

for (package in (.packages()) ) {
 print(paste("Package", package, "version", package.version(package)))
}
```

Load the data from the Sb vs SB allele specific comparison for both North American and South American populations, extract read counts.

```{r load_data, message = FALSE}
load("input/dds_Sb_vs_SB_north_america.RData")
load("input/dds_Sb_vs_SB_south_america.RData")

#Get normalised counts
dds_north_america <- estimateSizeFactors(dds_Bb_DE)
dds_south_america <- estimateSizeFactors(dds_deg_ar)

counts_north_america <- counts(dds_north_america, normalized = TRUE)
counts_south_america <- counts(dds_south_america, normalized = TRUE)
```

How many genes do both datasets have in common?

```{r genes_in_common, message = FALSE}
genes_south_america <- row.names(counts_south_america)
genes_north_america <- row.names(counts_north_america)

genes_both <- genes_north_america[genes_north_america %in% genes_south_america]

#Number of genes present in both datsets:
length(genes_both)
```

123 out of 125 genes found with fixed differences in South America are also present in North America

##Plot heatmap

Heatmap with raw read counts
```{r plot_heatmap_read_counts, message = FALSE}
#Parse the dataset to make it ggplot friendly
samples_north_america <- colnames(counts_north_america)
parsed_counts_north_america <- data.frame(c(counts_north_america),
                                          rep(samples_north_america, each = nrow(counts_north_america)),
                                          rep(genes_north_america, ncol(counts_north_america)),
                                          rep("north_america", (nrow(counts_north_america) * ncol(counts_north_america)))
                                          )
colnames(parsed_counts_north_america) <- c("counts", "sample", "gene", "population")

samples_south_america <- colnames(counts_south_america)
parsed_counts_south_america <- data.frame(c(counts_south_america),
                                          rep(samples_south_america, each = nrow(counts_south_america)),
                                          rep(genes_south_america, ncol(counts_south_america)),
                                          rep("south_america", (nrow(counts_south_america) * ncol(counts_south_america)))
                                          )
colnames(parsed_counts_south_america) <- c("counts", "sample", "gene", "population")


#Merge both datasets
parsed_counts_all <- rbind(parsed_counts_north_america, parsed_counts_south_america)
parsed_counts_all$allele <- gsub(x = parsed_counts_all$sample, pattern = ".+_", replacement = "")
parsed_counts_all$allele <- ifelse(parsed_counts_all$allele == "B" , "bigB", "littleb")
parsed_counts_all <- parsed_counts_all[order(parsed_counts_all$allele, parsed_counts_all$sample), ]

#Plot heatmap
ggplot(parsed_counts_all, aes(x = sample, y = gene)) + geom_tile(aes(fill = log2(counts + 1)), colour = "grey50") +
  facet_grid(. ~ population, scales = "free", space = "free") +
  scale_fill_gradient(low = "yellow" , high = "blue", name="log2 Estimated \nnormalized counts") +
  theme_bw() + theme( panel.grid.major=element_blank()) + theme(axis.text.x=element_blank(), axis.ticks.x=element_blank()) +
  theme(axis.title.x = element_blank(), axis.title.y = element_blank()) +
  theme(strip.text.x = element_text(size=20)) + theme(legend.title = element_text(size=15, face="bold"),
                                                      legend.text = element_text(size=13, face="bold"))

```


Heatmap with logarithm of the read count ratios between variants for North American populations
```{r plot_heatmap_ratio_north_america, message = FALSE}
#Parse the dataset to make it ggplot friendly
#Separate big B and little b samples
samples_north_america_littleb <- grep(x = samples_north_america, pattern = "_b", value = TRUE)
samples_north_america_bigB <- grep(x = samples_north_america, pattern = "_B", value = TRUE)
samples_north_america_no_allele <- unique(gsub(x = samples_north_america, pattern = "_[Bb]", replacement = ""))

counts_north_america_littleb <- counts_north_america[, samples_north_america_littleb]
counts_north_america_bigB <- counts_north_america[, samples_north_america_bigB]


parsed_counts_north_america_ratio <- data.frame(c(counts_north_america_littleb),
                                                c(counts_north_america_bigB),
                                                rep(samples_north_america_no_allele, each = nrow(counts_north_america)),
                                                rep(genes_north_america, length(samples_north_america_no_allele)),
                                                rep("north_america", (nrow(counts_north_america) * length(samples_north_america_no_allele)))
                                                )
colnames(parsed_counts_north_america_ratio) <- c("littleb", "bigB", "sample", "gene", "population")

parsed_counts_north_america_ratio$total_counts <- parsed_counts_north_america_ratio$littleb + parsed_counts_north_america_ratio$bigB
parsed_counts_north_america_ratio$lfc <- log2((parsed_counts_north_america_ratio$bigB +1) / (parsed_counts_north_america_ratio$littleb +1))

#Order the genes by mean read count
mean_reads_north_america <- ddply(parsed_counts_north_america_ratio, .(gene), summarize, mean =
                                  mean(total_counts))
mean_reads_north_america <- mean_reads_north_america[order(mean_reads_north_america$mean), ]
gene_order_north_america <- rep(mean_reads_north_america$gene,
                                length(unique(samples_north_america_no_allele)))

#Use the sorted genes to sort the parsed counts
parsed_counts_north_america_ratio$gene <- factor(parsed_counts_north_america_ratio$gene, levels = mean_reads_north_america$gene)

#Plot heatmap for North America
ggplot(parsed_counts_north_america_ratio, aes(x = sample, y = gene)) + geom_tile(aes(fill = lfc), colour = "grey50") +
       scale_fill_gradient(low = "yellow" , high = "blue", name="lfc") +
       theme_bw() + theme( panel.grid.major=element_blank()) + theme(axis.text.x=element_blank(), axis.ticks.x=element_blank()) +
       theme(axis.title.x = element_blank(), axis.title.y = element_blank()) +
       theme(strip.text.y = element_text(size=.5)) + theme(legend.title = element_text(size=15, face="bold"),
                                                           legend.text = element_text(size=13, face="bold"))
```

Same plot for genes present only in both populations
```{r plot_heatmap_ratio_north_america_genes_both, message = FALSE}

parsed_counts_north_america_ratio_subset <- parsed_counts_north_america_ratio[parsed_counts_north_america_ratio$gene %in% genes_both, ]

ggplot(parsed_counts_north_america_ratio_subset, aes(x = sample, y = gene)) + geom_tile(aes(fill = lfc), colour = "grey50") +
       scale_fill_gradient(low = "yellow" , high = "blue", name="lfc") +
       theme_bw() + theme( panel.grid.major=element_blank()) + theme(axis.text.x=element_blank(), axis.ticks.x=element_blank()) +
       theme(axis.title.x = element_blank(), axis.title.y = element_blank()) +
       theme(strip.text.y = element_text(size=.5)) + theme(legend.title = element_text(size=15, face="bold"),
                                                           legend.text = element_text(size=13, face="bold"))

```

Comparison of mean read counts between South American and North American populations
```{r compare_mean_read_counts, message = FALSE}
#Parse the dataset to make it ggplot friendly
#Separate big B and little b samples
samples_south_america_littleb <- grep(x = samples_south_america, pattern = "_b", value = TRUE)
samples_south_america_bigB <- grep(x = samples_south_america, pattern = "_B", value = TRUE)
samples_south_america_no_allele <- unique(gsub(x = samples_south_america, pattern = "_[Bb]", replacement = ""))

counts_south_america_littleb <- counts_south_america[, samples_south_america_littleb]
counts_south_america_bigB <- counts_south_america[, samples_south_america_bigB]

parsed_counts_south_america_ratio <- data.frame(c(counts_south_america_littleb),
                                                c(counts_south_america_bigB),
                                                rep(samples_south_america_no_allele,
                                                    each =nrow(counts_south_america)),
                                                rep(genes_south_america,
                                                    length(samples_south_america_no_allele)),
                                                rep("south_america",
                                                    (nrow(counts_south_america) * length(samples_south_america_no_allele)))
                                                )
colnames(parsed_counts_south_america_ratio) <- c("littleb", "bigB", "sample", "gene", "population")

parsed_counts_south_america_ratio$total_counts <- parsed_counts_south_america_ratio$littleb + parsed_counts_south_america_ratio$bigB
parsed_counts_south_america_ratio$lfc <- log2((parsed_counts_south_america_ratio$bigB +1) / (parsed_counts_south_america_ratio$littleb +1))

#Get mean read count for South American samples
mean_reads_south_america <- ddply(parsed_counts_south_america_ratio, .(gene), summarize, mean =
                                  mean(total_counts))
mean_reads_south_america <- mean_reads_south_america[order(mean_reads_south_america$mean), ]

#Get the list of genes in North America that overlap with South America
mean_reads_all <- merge(mean_reads_north_america, mean_reads_south_america, by = "gene")
colnames(mean_reads_all) <- c("gene", "mean_na", "mean_sa")

#Plot means per population
plot(log(mean_reads_all$mean_na),
     log(mean_reads_all$mean_sa))

```

Mean read counts across populations is roughly similar. For next plots, the order for the North American dataset will be kept.
Heatmap with logarithm of the read count ratios between variants for South American populations
```{r plot_heatmap_ratio_south_america, message = FALSE}

#Use the sorted genes from the North America dataset to sort the parsed counts
mean_reads_all_sorted <- mean_reads_all[order(mean_reads_all$mean_na), ]
parsed_counts_south_america_ratio$gene <- factor(parsed_counts_south_america_ratio$gene, levels = mean_reads_all_sorted$gene)

#Add body part as a factor
parsed_counts_south_america_ratio$body_part <- gsub(x = parsed_counts_south_america_ratio$sample,
                                                    pattern = "([0-9]+[BC]?)([A-Z]+)", replacement = "\\2")

#Plot heatmap for South America
ggplot(parsed_counts_south_america_ratio, aes(x = sample, y = gene)) + geom_tile(aes(fill = lfc), colour = "grey50") +
       facet_grid(. ~ body_part, scales = "free") +
       scale_fill_gradient(low = "yellow" , high = "blue", name="log2 Estimated \nnormalized counts") +
       theme_bw() + theme( panel.grid.major=element_blank()) + theme(axis.text.x=element_blank(), axis.ticks.x=element_blank()) +
       theme(axis.title.x = element_blank(), axis.title.y = element_blank()) +
       theme(strip.text.x = element_text(size=20)) + theme(legend.title = element_text(size=15, face="bold"),
                                                           legend.text = element_text(size=13, face="bold"))
```

Plot heatmap for genes present only in both datasets
```{r plot_heatmap_ratio_both, message = FALSE}
#Ensure the datasets for both populations have the same genes.
parsed_counts_north_america_ratio$body_part <- rep("whole_body_queen", nrow(parsed_counts_north_america_ratio))
parsed_counts_all_ratios_common_genes <- rbind(parsed_counts_north_america_ratio, parsed_counts_south_america_ratio)
parsed_counts_all_ratios_common_genes <- parsed_counts_all_ratios_common_genes[parsed_counts_all_ratios_common_genes$gene %in% genes_both, ]

ggplot(parsed_counts_all_ratios_common_genes, aes(x = sample, y = gene)) + geom_tile(aes(fill = lfc), colour = "grey50") +
       facet_grid(. ~ body_part, scales = "free", space = "free") +
       scale_fill_gradient(low = "yellow" , high = "blue", name="log2 Estimated \nnormalized counts") +
       theme_bw() + theme( panel.grid.major=element_blank()) + theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
       theme(axis.title.x = element_blank(), axis.title.y = element_blank()) +
       theme(strip.text.x = element_text(size=20)) + theme(legend.title = element_text(size=15, face="bold"),
                                                           legend.text = element_text(size=13, face="bold"))
```


Plot heatmap by read counts for genes present in both populations

```{r plot_heatmap_read_counts_compare_populations, message = FALSE}
#Ensure the datasets for both populations have the same genes.
parsed_counts_all_common_genes <- parsed_counts_all[parsed_counts_all$gene %in% genes_both, ]

#Add a body_part factor
parsed_counts_all_common_genes$body_part <- gsub(x = parsed_counts_all_common_genes$sample,
                                                 pattern = "([0-9]+[BC]?Q?)([A-Z])", replacement = "\\2")
parsed_counts_all_common_genes$body_part <- gsub(x = parsed_counts_all_common_genes$body_part,
                                                 pattern = "[0-9]?_[Bb]", replacement = "")

#Make sure the heatmap is plotted ordered by allele
levels_bigB <- grep("_B", levels(parsed_counts_all_common_genes$sample), value = TRUE)
levels_littleb <- grep("_b", levels(parsed_counts_all_common_genes$sample), value = TRUE)
all_levels <- c(levels_bigB, levels_littleb)
parsed_counts_all_common_genes$sample <- factor(parsed_counts_all_common_genes$sample,
                                                levels = all_levels)

#Get the names for the x lables in the heatmap (SB vs Sb)
allele_names_x_axis <- gsub(x = all_levels, pattern = ".+_", replacement = "")

ggplot(parsed_counts_all_common_genes, aes(x = sample, y = gene)) + geom_tile(aes(fill = log2(counts + 1)), colour = "grey50") +
  facet_grid(. ~ body_part, scales = "free", space = "free") +
  scale_fill_gradient(low = "yellow" , high = "blue", name="log2 Estimated \nnormalized counts") +
  theme_bw() + theme( panel.grid.major=element_blank()) + theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  theme(axis.title.x = element_blank(), axis.title.y = element_blank()) +
  theme(strip.text.x = element_text(size=20)) + theme(legend.title = element_text(size=15, face="bold"),
                                                      legend.text = element_text(size=13, face="bold"))

```

Plot each population individually, ordering genes by p value
For South America
```{r plot_heatmap_read_counts_south_america, message = FALSE}

#Get p values for all genes
results_north_america <- results(dds_Bb_DE)
results_south_america <- results(dds_deg_ar)

#Order genes by pvalues
results_north_america <- results_north_america[order(results_north_america$pvalue, decreasing = TRUE), ]
results_south_america <- results_south_america[order(results_south_america$pvalue, decreasing = TRUE), ]

#Get the names for the x lables in the heatmap (SB vs Sb)
gene_order_pval_north_america <- rownames(results_north_america)
gene_order_pval_south_america <- rownames(results_south_america)

#Add body_part as a factor
parsed_counts_south_america$body_part <- gsub(x = parsed_counts_south_america$sample,
                                              pattern = "([0-9]+[BC]?Q?)([A-Z])", replacement = "\\2")

parsed_counts_south_america$body_part <- gsub(x = parsed_counts_south_america$body_part,
                                              pattern = "_.+", replacement = "")

#Get the right order for genes and samples
parsed_counts_south_america$sample <- factor(parsed_counts_south_america$sample,
                                             levels = all_levels)

parsed_counts_south_america$gene <- factor(parsed_counts_south_america$gene,
                                           levels = gene_order_pval_south_america)

#Plot the heatmap
ggplot(parsed_counts_south_america, aes(x = sample, y = gene)) + geom_tile(aes(fill = log2(counts + 1)), colour = "grey50") +
       facet_grid(. ~ body_part, scales = "free", space = "free") +
       scale_fill_gradient(low = "yellow" , high = "blue", name="log2 Estimated \nnormalized counts") +
       theme_bw() + theme( panel.grid.major=element_blank()) + theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
       theme(axis.title.x = element_blank(), axis.title.y = element_blank()) +
       theme(strip.text.x = element_text(size=20)) + theme(legend.title = element_text(size=15, face="bold"),
                                                      legend.text = element_text(size=13, face="bold"))


```

For North America
```{r plot_heatmap_read_counts_north_america, message = FALSE}
#Get the right order for genes and samples
parsed_counts_north_america$sample <- factor(parsed_counts_north_america$sample,
                                             levels = all_levels)

parsed_counts_north_america$gene <- factor(parsed_counts_north_america$gene,
                                           levels = gene_order_pval_north_america)

#Plot the heatmap
ggplot(parsed_counts_north_america, aes(x = sample, y = gene)) + geom_tile(aes(fill = log2(counts + 1)), colour = "grey50") +
       scale_fill_gradient(low = "yellow" , high = "blue", name="log2 Estimated \nnormalized counts") +
       theme_bw() + theme( panel.grid.major=element_blank()) + theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
       theme(axis.title.x = element_blank(), axis.title.y = element_blank()) +
       theme(strip.text.x = element_text(size=20)) + theme(legend.title = element_text(size=15, face="bold"),
                                                      legend.text = element_text(size=13, face="bold"))



```

## Fisher's combined probability test

Because it is difficult to compare North and South American p values directly, the results of both tests will be fused together using (Fisher's method)<https://en.wikipedia.org/wiki/Fisher%27s_method>. This method will test whether the combined p value of both analyses is significant.

```{r fisher_combined_test, message = FALSE}
#Obtain p values for genes present only in both datasets
results_north_america_subset <- results_north_america[genes_both, ]
results_south_america_subset <- results_south_america[genes_both, ]

#Get a vector of X2 values based on the combination of p values of both datasets
combined_ps <- -2 * log((results_north_america_subset$pvalue + results_north_america_subset$pvalue))

#Obtain new p values based on X2 distribution
new_ps <- pchisq(combined_ps, df = 4, lower.tail = FALSE)

#Adjust p values using the Benjamini and Hochberg method
new_ps_adjust <- p.adjust(new_ps, method = "BH")

#Because the genes where in the same order, the new p values can be named directly:
names(new_ps_adjust) <- genes_both

#Retrieve significant genes from the combined p values
sig_genes_both <- names(new_ps_adjust[new_ps_adjust < 0.05])
    ```

Re-run the DESeq2 analysis for the allele:body_part interaction, but enriched with genes differentially expressed in both populations.

```{r allele:body_part_interaction_subset, message = FALSE}
#Retrieve colData from the DESeq dataset and subset for genes of interest
colData_subset <- colData(dds_deg_ar)

#Subset dataset to include only genes of interest
raw_counts_south_america_subset <- counts(dds_deg_ar, normalized = FALSE)[sig_genes_both, ]

#Re-run DESeq2
dds_ar_interaction_subset <- DESeqDataSetFromMatrix(countData = raw_counts_south_america_subset, colData = colData_subset, design = ~ colony + body_part + allele + allele:body_part)
#Perform the analysis replacing the sizeFactors:
sizeFactors(dds_ar_interaction_subset) <- rep(1, ncol(raw_counts_south_america_subset))
#If using test = LRT, deseq2 performs a test for detecting DE loci, using first a likelihood ratio test (ANOVA-like). This test compares the model ~body_part+allele against a
#reduced model, ~body_part. The p values indicate which genes are significantly DE accross ALL levels of allele. Normalisation
#has been done only by boy_part, as ASE data comes always from the same sample. If test is not specified it runs the usual Wald test for the last element in the design formula (body_part:allele)
dds_deg_ar_interaction_subset <- DESeq(dds_ar_interaction_subset, test = "LRT", reduced = ~ colony + body_part + allele)
res_interaction_subset <- results(dds_deg_ar_interaction_subset)

#Plot the genes of interest for each allele
subset_parsed_counts_south_america <- parsed_counts_south_america[parsed_counts_south_america$gene %in% sig_genes_both, ]

subset_parsed_counts_south_america$allele <- gsub(x = subset_parsed_counts_south_america$sample,
                                                  pattern = ".+_",
                                                  replacement = "")

ggplot(subset_parsed_counts_south_america, aes(x = allele, y = log(counts +1), fill = body_part)) + geom_boxplot() +
  facet_grid(. ~ gene)
```

## lmer for NA vs SA genes, are there significant common patterns in expression differences?

The Fisher's method for p values combination only takes the p values, and is blind to any other information like, for instance, whether the direction of expression is the same in both comparisons. 
To overcome this, I will run an lme that will account for tissue and population, a combined test that cannot be run on DESeq. The output should tell us which genes are DE between variants across populations.

```{r lmer_na_vs_sa, message = FALSE}

#Normalise the LFCs between factors using Z scores (center in normal distribution)
#Calculate the mean and sd lfcs per body part/population
agg_means <- aggregate(lfc ~ body_part, data = parsed_counts_all_ratios_common_genes, mean)
agg_sd <- aggregate(lfc ~ body_part, data = parsed_counts_all_ratios_common_genes, sd)

for_z_scores <- merge(agg_means, agg_sd, by = "body_part")
colnames(for_z_scores) <- c("body_part", "mean_body_part", "sd_body_part")

#Add means and sd to the main dataset
parsed_counts_all_ratios_common_genes <- merge(for_z_scores, parsed_counts_all_ratios_common_genes)

#Get the normalised lfcs
parsed_counts_all_ratios_common_genes$lfc_norm <- (parsed_counts_all_ratios_common_genes$lfc - parsed_counts_all_ratios_common_genes$mean_body_part) /
                                                   parsed_counts_all_ratios_common_genes$sd_body_part

#Get an estimate of the relationship between read counts and its effect on LFC variance
results_north_america_df <- as.data.frame(results_north_america)
plot(log(results_north_america_df$baseMean), 1 / (results_north_america_df$lfcSE)^2)
#Looks quadratic?
yvals <- 1 / (results_north_america_df$lfcSE) ^ 2 
xvals <- log(results_north_america_df$baseMean)
se_reads <- lm(yvals ~ xvals + I(xvals ^ 2) + I(xvals ^ 3))

plot(xvals, yvals)
re_order <- order(xvals) 
lines(xvals[re_order], fitted(se_reads)[re_order])

#Pretty good fit. Use the model between log of read counts and variance to caculate the weights for the lfcs estimates 
weights_model <- predict(se_reads,
                         data.frame(xvals = log(parsed_counts_all_ratios_common_genes$total_counts + 0.1)))
#Make sure there are no 0s or negative values for the model (otherwise it won't converge). This removes replicates for genes that
#convey no information anyway
for_model <- parsed_counts_all_ratios_common_genes
for_model$weights <- weights_model 
for_model <- for_model[for_model$weights > 0, ]

#Run model
lme_populations <- lmer(lfc_norm ~ 0 + gene * population + (1|body_part:gene), data = for_model,
                        weights = weights)

# extract coefficients
coefs <- data.frame(coef(summary(lme_populations)))
#Adjust p values
coefs$padj <- p.adjust(coefs$Pr...t.., method = "BH")
#Significant genes (only those with main gene effects are considered)
sig_genes_both_lme <- coefs[coefs$padj < 0.05, ]

#Get the names of the significant genes from the model (main effects only)
sig_genes_both_lme <- rownames(sig_genes_both_lme)
#Get main effects only
sig_genes_both_lme <- sig_genes_both_lme[!grepl(pattern = ":population", x = sig_genes_both_lme)]
#Clean the names
sig_genes_both_lme <- gsub(pattern = "gene", replacement = "", x = sig_genes_both_lme)
  
```

```{r boxplot_ratios, message = FALSE, warning = FALSE}
#Calculate median read counts per group (gene and body part)
agg_medians <- aggregate(total_counts ~ gene + body_part, data = parsed_counts_all_ratios_common_genes, median)
agg_medians <- agg_medians[order(agg_medians$gene, agg_medians$body_part), ]
colnames(agg_medians) <- c("gene", "body_part", "medians")

#Add medians to the main dataset
parsed_counts_all_ratios_common_genes <- merge(agg_medians, parsed_counts_all_ratios_common_genes)

#Get the order of all genes by position in the supergene
#Load the annotation for the gnG assembly of the Solenopsis invicta reference genome.
si_ann <- makeTxDbFromGFF(file = "input/GCF_000188075.1_Si_gnG_genomic.gff",
                          format="gff3")

#Generate a table with the gene names and its position in the reference
gene_ids <- keys(si_ann, "GENEID")
gene_positions <- AnnotationDbi::select(x = si_ann,
                                        keys = gene_ids,
                                        columns = c("GENEID","TXCHROM", "TXSTART", "TXEND"),
                                        keytype = "GENEID")

colnames(gene_positions) <- c("gene","contig", "start", "end")

#Get only the minimum start position in the scaffolds for all the transcripts per gene
gene_positions_min <- aggregate(x = gene_positions$start, by = list(gene_positions$contig, gene_positions$gene), FUN = min)
names(gene_positions_min) <- c("contig", "gene", "start_position")

#Select genes only in the analysis
gene_positions_min <- gene_positions_min[gene_positions_min$gene %in% unique(parsed_counts_all_ratios_common_genes$gene), ]

#Sort by start position per contig
gene_positions_min <- gene_positions_min[order(gene_positions_min$contig, gene_positions_min$start_position), ]

#Mark the genes with significant DE in the different populations and in both
sig_genes_na <- rownames(results_north_america_subset[which(results_north_america_subset$padj < 0.05), ])

sig_genes_sa <- rownames(results_south_america_subset[which(results_south_america_subset$padj < 0.05), ])

parsed_counts_all_ratios_common_genes$gene <- as.character(parsed_counts_all_ratios_common_genes$gene)

parsed_counts_all_ratios_common_genes$gene[parsed_counts_all_ratios_common_genes$gene %in% sig_genes_both_lme] <- paste0(parsed_counts_all_ratios_common_genes$gene[parsed_counts_all_ratios_common_genes$gene %in% sig_genes_both_lme], "*+")

parsed_counts_all_ratios_common_genes$gene[parsed_counts_all_ratios_common_genes$gene %in% sig_genes_na] <- paste0(parsed_counts_all_ratios_common_genes$gene[parsed_counts_all_ratios_common_genes$gene %in% sig_genes_na], "*")

parsed_counts_all_ratios_common_genes$gene[parsed_counts_all_ratios_common_genes$gene %in% sig_genes_sa] <- paste0(parsed_counts_all_ratios_common_genes$gene[parsed_counts_all_ratios_common_genes$gene %in% sig_genes_sa], "+")

#Mark in both datasets
gene_positions_min$gene[gene_positions_min$gene %in% sig_genes_both_lme] <- paste0(gene_positions_min$gene[gene_positions_min$gene %in% sig_genes_both_lme], "*+")

gene_positions_min$gene[gene_positions_min$gene %in% sig_genes_na] <- paste0(gene_positions_min$gene[gene_positions_min$gene %in% sig_genes_na], "*")

gene_positions_min$gene[gene_positions_min$gene %in% sig_genes_sa] <- paste0(gene_positions_min$gene[gene_positions_min$gene %in% sig_genes_sa], "+")

#Sort the levels in the main dataframe
parsed_counts_all_ratios_common_genes$gene <- factor(parsed_counts_all_ratios_common_genes$gene, 
                                                     levels = gene_positions_min$gene)

facet_labels <- c(QA = "South America\nQueen Abdomen", QH = "South America\nQueen Head", 
                  QT = "South America\nQueen Thorax", W = "South America\nWorker Whole",
                  whole_body_queen = "North America\nQueen Whole")
ggplot(data = parsed_counts_all_ratios_common_genes, aes(x = gene, y = lfc)) + geom_boxplot(aes(fill = log(medians + 1)),
                                                                                            outlier.size = 0.5) + 
  scale_fill_gradient(name = "Logarithm of median\nread counts +1", low = hcl(15,100,75), high = hcl(195,100,75)) +
  facet_grid(rows = vars(body_part), scales = "free", labeller = labeller(body_part = facet_labels)) +
  geom_hline(yintercept = 0, color = "darkblue") + geom_hline(yintercept = -1, color = "darkblue", linetype = "dashed") + 
  geom_hline(yintercept = 1, color = "darkblue", linetype = "dashed") + labs(y = "Log2 fold differences") + 
  theme_bw() + theme(panel.grid.minor = element_blank()) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 10), axis.title.x = element_blank(),
        axis.text.y = element_text(vjust = 0.5, size = 12), axis.title.y = element_text(size = 14, face = "bold"),
        strip.text.y= element_text(size = 12, angle = 45), legend.title = element_text(size=13, face="bold"),
        legend.text = element_text(size=12))
```

Interesting patterns to note emerging from this graph:

* (LOC105202834)<https://www.ncbi.nlm.nih.gov/gene/?term=LOC105202834> and (LOC105202818)<https://www.ncbi.nlm.nih.gov/gene/?term=LOC105202818> differences between SB and Sb are exactly the same in all SA samples. Both genes are annotated as "cytochrome P450 4C1" Presumably the reads here mapped to the same place. Is it duplicated in NA only (if not an artifact)?

* (LOC105193134)<https://www.ncbi.nlm.nih.gov/gene/?term=LOC105193134> was picked up as significant with the Fisher combination method for p values, but the LFCs estimates looked all over the place. In this graph they are clearly above 0 in every comparison and almost always above 1 LFC (2 fold difference between SB and Sb).

```{r save_boxplot, message = FALSE, warning = FALSE}

p1 <- ggplot(data = parsed_counts_all_ratios_common_genes, aes(x = gene, y = lfc)) + geom_boxplot(aes(fill = log(medians + 1)),
                                                                                                      outlier.size = 0.5) + 
              scale_fill_gradient(name = "Logarithm of median\nread counts +1", low = hcl(15,100,75), high = hcl(195,100,75)) +
              facet_grid(rows = vars(body_part), scales = "free", labeller = labeller(body_part = facet_labels)) +
              geom_hline(yintercept = 0, color = "darkblue") + geom_hline(yintercept = -1, color = "darkblue", linetype = "dashed") + 
              geom_hline(yintercept = 1, color = "darkblue", linetype = "dashed") + labs(y = "Log2 fold differences") +
              theme_bw() + theme(panel.grid.minor = element_blank()) +
              theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 10), axis.title.x = element_blank(),
                    axis.text.y = element_text(vjust = 0.5, size = 12), axis.title.y = element_text(size = 14, face = "bold"),
                    strip.text.y= element_text(size = 12, angle = 45), legend.title = element_text(size=13, face="bold"),
                    legend.text = element_text(size=12))

ggsave(filename = "results/boxplot_lfcs_common.pdf", plot = p1, device = "pdf", height = 30, width = 60, units = "cm")
```

Same plot with North American populations only

```{r boxplot_ratios_north_america, message = FALSE, warning = FALSE}
#Calculate median read counts per group (gene and body part)
agg_medians_north_america <- aggregate(total_counts ~ gene + body_part, data = parsed_counts_north_america_ratio, median)
agg_medians_north_america <- agg_medians_north_america[order(agg_medians_north_america$gene, agg_medians_north_america$body_part), ]
colnames(agg_medians_north_america) <- c("gene", "body_part", "medians")

#Add medians to the main dataset
parsed_counts_north_america_ratio <- merge(agg_medians_north_america, parsed_counts_north_america_ratio)

#Get only the minimum start position in the scaffolds for all the transcripts per gene
gene_positions_min_north_america <- aggregate(x = gene_positions$start, by = list(gene_positions$contig, gene_positions$gene), FUN = min)
names(gene_positions_min_north_america) <- c("contig", "gene", "start_position")

#Select genes only in the analysis
gene_positions_min_north_america <- gene_positions_min_north_america[gene_positions_min_north_america$gene %in% unique(parsed_counts_north_america_ratio$gene), ]

#Sort by start position per contig
gene_positions_min_north_america <- gene_positions_min_north_america[order(gene_positions_min_north_america$contig, gene_positions_min_north_america$start_position), ]

#Mark the genes with significant DE in the different populations and in both
sig_genes_na_all <- rownames(results_north_america[which(results_north_america$padj < 0.05), ])
gene_positions_min_north_america$gene[gene_positions_min_north_america$gene %in% sig_genes_na_all] <- paste0(gene_positions_min_north_america$gene[gene_positions_min_north_america$gene %in% sig_genes_na_all], "*")
parsed_counts_north_america_ratio$gene <- as.character(parsed_counts_north_america_ratio$gene)
parsed_counts_north_america_ratio$gene[parsed_counts_north_america_ratio$gene %in% sig_genes_na_all] <- paste0(parsed_counts_north_america_ratio$gene[parsed_counts_north_america_ratio$gene %in% sig_genes_na_all], "*")

#Sort the levels in the main dataframe
parsed_counts_north_america_ratio$gene <- factor(parsed_counts_north_america_ratio$gene, 
                                                 levels = gene_positions_min_north_america$gene)

p2 <- ggplot(data = parsed_counts_north_america_ratio, aes(x = gene, y = lfc)) + geom_boxplot(aes(fill = log(medians + 1)),
                                                                                              outlier.size = 0.5) + 
              scale_fill_gradient(name = "Logarithm of median\nread counts +1", low = hcl(15,100,75), high = hcl(195,100,75)) +
              geom_hline(yintercept = 0, color = "darkblue") + geom_hline(yintercept = -1, color = "darkblue", linetype = "dashed") + 
              geom_hline(yintercept = 1, color = "darkblue", linetype = "dashed") + labs(y = "Log2 fold differences") +
              #geom_rect(aes(xmin = sig_genes_both[1], xmax = sig_genes_both[1], ymin = -8, ymax = 8),
              #          fill = "transparent", color = "red", size = 1.5) + 
              theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 5), axis.title.x = element_blank(),
                    axis.text.y = element_text(vjust = 0.5, size = 12), axis.title.y = element_text(size = 14, face = "bold"),
                    strip.text.y= element_text(size = 12, angle = 45), legend.title = element_text(size=13, face="bold"),
                    legend.text = element_text(size=12))

p2
#ggsave(filename = "results/boxplot_lfcs_north_america.png", plot = p2, device = "png", height = 60, width = 120, units = "cm")

```

Repeat boxplot in common with different format settings (width of the boxplot linked to significance)

```{r boxplot_ratios_common_format2, message = FALSE, warning = FALSE}

#Add a column specifying which genes are significant in which populations
parsed_counts_all_ratios_common_genes$sig <- rep ("no", nrow(parsed_counts_all_ratios_common_genes))
parsed_counts_all_ratios_common_genes[parsed_counts_all_ratios_common_genes$gene %in% paste0(sig_genes_sa, "+"), ]$sig <- "sa"
parsed_counts_all_ratios_common_genes[parsed_counts_all_ratios_common_genes$gene %in% paste0(sig_genes_both_lme, "*+"), ]$sig <- "both"
#Make sure that the label 'sa' only applies to genes in south american populations
parsed_counts_all_ratios_common_genes[parsed_counts_all_ratios_common_genes$sig == "sa" &
                                      parsed_counts_all_ratios_common_genes$population == "north_america", ]$sig <- "no"  

#Generate x axis only with significant genes

x_axis_genes <- rep("", length(levels(parsed_counts_all_ratios_common_genes$gene)))

pos_sig_genes <- c(which(levels(parsed_counts_all_ratios_common_genes$gene) %in% paste0(sig_genes_both_lme, "*+")), 
                   which(levels(parsed_counts_all_ratios_common_genes$gene) %in% paste0(sig_genes_sa, "+")))

pos_sig_genes <- sort(pos_sig_genes)
x_axis_genes_ids <- levels(parsed_counts_all_ratios_common_genes$gene)[pos_sig_genes]
x_axis_genes[pos_sig_genes] <- x_axis_genes_ids
x_axis_genes <- gsub(x = x_axis_genes, pattern = "[*+]", replacement = "")

#Plot the thing
p3 <- ggplot(data = parsed_counts_all_ratios_common_genes, aes(x = gene, y = lfc)) + geom_boxplot(varwidth = TRUE, aes(weight = log(medians + 1.1), 
                                                                                                                 fill = sig, color = sig), outlier.size = 0.5) +
        scale_fill_manual(values=c(hcl(15,100,75), "#999999", hcl(195,100,75))) +
        scale_color_manual(values=c(hcl(15,100,50), "grey40", hcl(195,100,50))) +
        facet_grid(rows = vars(body_part), scales = "free", labeller = labeller(body_part = facet_labels)) +
        geom_hline(yintercept = 0, color = "black") + geom_hline(yintercept = -1, color = "darkgrey", linetype = "dashed") + 
        geom_hline(yintercept = 1, color = "darkgrey", linetype = "dashed") +  
        theme_bw() + theme(panel.grid.minor = element_blank()) + scale_x_discrete(labels = x_axis_genes) +
        scale_y_continuous(breaks = seq(-10, 10, 2), expand = expand_scale(mult = 0)) +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 10), axis.title.x = element_blank(),
              axis.text.y = element_text(vjust = 0.5, size = 12), axis.title.y = element_blank(),
              strip.text.y= element_text(size = 12, angle = 45), legend.position = "none")
        

#Save the plot
ggsave(filename = "results/boxplot_lfcs_common_format2.pdf", plot = p3, device = "pdf", height = 210, width = 297, units = "mm")

```

Run the same format for the Noth American only plot

```{r boxplot_ratios_north_america_format2, message = FALSE, warning = FALSE}

#Add a column specifying which genes are significant 
parsed_counts_north_america_ratio$sig <- rep ("no", nrow(parsed_counts_north_america_ratio))
parsed_counts_north_america_ratio[parsed_counts_north_america_ratio$gene %in% paste0(sig_genes_na_all, "*"), ]$sig <- "na"

#Generate x axis only with significant genes
x_axis_genes <- rep("", length(levels(parsed_counts_north_america_ratio$gene)))

pos_sig_genes <- which(levels(parsed_counts_north_america_ratio$gene) %in% paste0(sig_genes_na_all, "*"))

pos_sig_genes <- sort(pos_sig_genes)
x_axis_genes_ids <- levels(parsed_counts_north_america_ratio$gene)[pos_sig_genes]
x_axis_genes[pos_sig_genes] <- x_axis_genes_ids
x_axis_genes <- gsub(x = x_axis_genes, pattern = "[*]", replacement = "")

#Plot the thing
p3 <- ggplot(data = parsed_counts_north_america_ratio, aes(x = gene, y = lfc)) + geom_boxplot(varwidth = TRUE, aes(weight = log(medians + 1.1), 
                                                                                                                 fill = sig, color = sig), outlier.size = 0.5) +
        scale_fill_manual(values=c("darkolivegreen3", "#999999")) +
        scale_color_manual(values=c("darkgreen", "grey40")) +
        geom_hline(yintercept = 0, color = "black") + geom_hline(yintercept = -1, color = "darkgrey", linetype = "dashed") + 
        geom_hline(yintercept = 1, color = "darkgrey", linetype = "dashed") +  
        theme_bw() + theme(panel.grid.minor = element_blank()) + scale_x_discrete(labels = x_axis_genes) +
        scale_y_continuous(breaks = seq(-10, 10, 2), expand = expand_scale(mult = 0)) +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 10), axis.title.x = element_blank(),
              axis.text.y = element_text(vjust = 0.5, size = 12), axis.title.y = element_blank(),
              strip.text.y= element_text(size = 12, angle = 45), legend.position = "none")
        

#Save the plot
ggsave(filename = "results/boxplot_lfcs_north_america_format2.pdf", plot = p3, device = "pdf", height = 210, width = 297, units = "mm")

```


Similar boxplot, but including only genes with significant differences

```{r boxplot_ratios_common_format2_sig_only, message = FALSE, warning = FALSE}

#Remove the + and * from the gene names, as they are no longer necessary
parsed_counts_sig_ratios_common_genes <- parsed_counts_all_ratios_common_genes
parsed_counts_sig_ratios_common_genes$gene <- gsub(pattern = "\\+|\\*",
                                                   replacement = "",
                                                   x = parsed_counts_all_ratios_common_genes$gene)

#Select only significant genes for the genes in common between SA and NA
sig_genes_plot <- unique(c(sig_genes_both_lme, sig_genes_sa))
parsed_counts_sig_ratios_common_genes <- subset(x = parsed_counts_sig_ratios_common_genes, 
                                                subset = parsed_counts_sig_ratios_common_genes$gene %in% sig_genes_plot)




#Include non-significant genes in North American populations


p4 <- ggplot(data = parsed_counts_sig_ratios_common_genes, aes(x = gene, y = lfc)) + geom_boxplot(varwidth = TRUE, aes(weight = log(medians + 1.1), 
                                                                                                                 fill = sig, color = sig), outlier.size = 0.5) +
        scale_fill_manual(values=c(hcl(15,100,75), "#999999", hcl(195,100,75))) +
        scale_color_manual(values=c(hcl(15,100,50), "grey40", hcl(195,100,50))) +
        facet_grid(rows = vars(body_part), labeller = labeller(body_part = facet_labels)) +
        geom_hline(yintercept = 0, color = "black") + geom_hline(yintercept = -1, color = "darkgrey", linetype = "dashed") + 
        geom_hline(yintercept = 1, color = "darkgrey", linetype = "dashed") +  
        theme_bw() + theme(panel.grid.minor = element_blank()) +
        scale_y_continuous(breaks = seq(-10, 10, 2), expand = expand_scale(mult = 0)) +
        theme(axis.text.x = element_text(angle = 45, vjust = 0.5, size = 18), axis.title.x = element_blank(),
              axis.text.y = element_text(vjust = 0.5, size = 12), axis.title.y = element_blank(),
              strip.text.y= element_text(size = 12, angle = 45), legend.position = "none")
        

#Save the plot
ggsave(filename = "results/boxplot_lfcs_common_sig_only_format2.pdf", plot = p4, device = "pdf", height = 210, width = 297, units = "mm")

```

Boxplot with significant genes in both SA and NA, comparing lfcs per gene across populations/body parts

```{r boxplot_ratios_common_sig_only_per_gene, message = FALSE, warning = FALSE}

#Remove LOC from the gene IDs
parsed_counts_sig_ratios_common_genes$gene <- gsub(pattern = "LOC",
                                                   replacement = "",
                                                   x = parsed_counts_sig_ratios_common_genes$gene)


p5 <- ggplot(data = parsed_counts_sig_ratios_common_genes, aes(x = body_part, y = lfc)) + geom_boxplot(varwidth = TRUE, aes(weight = log(medians + 1.1), 
                                                                                                                 fill = sig, color = sig), outlier.size = 0.5) +
               scale_fill_manual(values=c(hcl(15,100,75), "#999999", hcl(195,100,75)),
                          breaks = c("both", "sa", "no"), 
                          labels = c("North and South\nAmerica", "South America\nonly", "No"),
                          name = "Significant differences\nbetween variants") +
        scale_color_manual(values=c(hcl(15,100,50), "grey40", hcl(195,100,50)),
                          breaks = c("both", "sa", "no"), 
                          labels = c("North and South\nAmerica", "South America\nonly", "No"),
                          name = "Significant differences\nbetween variants") +
        facet_grid(. ~ gene) +
        geom_hline(yintercept = 0, color = "black") + geom_hline(yintercept = -1, color = "darkgrey", linetype = "dashed") + 
        geom_hline(yintercept = 1, color = "darkgrey", linetype = "dashed") +  
        theme_bw() + theme(panel.grid.minor = element_blank()) +
        scale_y_continuous(breaks = seq(-10, 10, 2)) + #, expand = expand_scale(mult = 0)) +
        scale_x_discrete(labels=c("QA" = "1", "QH" = "2", "QT" = "3", "W" = "4", "whole_body_queen" = "5")) + 
        theme(axis.text.x = element_text(vjust = 0.5, size = 12), axis.title.x = element_blank(),
              axis.text.y = element_text(vjust = 0.5, size = 12), axis.title.y = element_blank(),
              strip.text.x= element_text(size = 9), legend.title = element_text(size = 15),
              legend.text = element_text(size = 12))
        
p6 <- ggplot(data = parsed_counts_sig_ratios_common_genes, aes(x = body_part, y = lfc)) + geom_boxplot(varwidth = TRUE, aes(weight = log(medians + 1.1), 
                                                                                                                 fill = sig, color = sig), outlier.size = 0.5) +
        scale_fill_manual(values=c(hcl(15,100,75), "#999999", hcl(195,100,75)),
                          breaks = c("both", "sa", "no"), 
                          labels = c("North and South\nAmerica", "South America\nonly", "No"),
                          name = "Significant differences\nbetween variants") +
        scale_color_manual(values=c(hcl(15,100,50), "grey40", hcl(195,100,50)),
                          breaks = c("both", "sa", "no"), 
                          labels = c("North and South\nAmerica", "South America\nonly", "No"),
                          name = "Significant differences\nbetween variants") +
        facet_grid(. ~ gene) +
        geom_hline(yintercept = 0, color = "black") + geom_hline(yintercept = -1, color = "darkgrey", linetype = "dashed") + 
        geom_hline(yintercept = 1, color = "darkgrey", linetype = "dashed") +  
        theme_bw() + theme(panel.grid.minor = element_blank()) +
        scale_y_continuous(breaks = seq(-10, 10, 2)) + #, expand = expand_scale(mult = 0)) +
        scale_x_discrete(labels=c("QA" = "1", "QH" = "2", "QT" = "3", "W" = "4", "whole_body_queen" = "5")) + 
        theme(axis.text.x = element_text(vjust = 0.5, size = 12), axis.title.x = element_blank(),
              axis.text.y = element_text(vjust = 0.5, size = 12), axis.title.y = element_blank(),
              strip.text.x= element_text(size = 9), legend.position = "none")
#Save the plot
ggsave(filename = "results/boxplot_lfcs_common_sig_only_per_gene.pdf", plot = p6, device = "pdf", height = 150, width = 297, units = "mm")
ggsave(filename = "results/boxplot_lfcs_common_sig_only_per_gene_legend.pdf", plot = p5, device = "pdf", height = 150, width = 297, units = "mm")

```

Similar boxplot putting the focus on body parts

```{r boxplot_ratios_common_sig_only_per_gene_body_part, message = FALSE, warning = FALSE}

fill_colors <- c(rgb(152,78,163, maxColorValue = 255), rgb(55,126,184, maxColorValue = 255),
                 rgb(77,175,74, maxColorValue = 255), rgb(228,26,28, maxColorValue = 255),
                 rgb(255,127,0, maxColorValue = 255))
contour_colors <- c(rgb(106, 55, 114, maxColorValue = 255), rgb(39, 88, 129, maxColorValue = 255),
                 rgb(54, 122, 52, maxColorValue = 255), rgb(160,18,18, maxColorValue = 255),
                 rgb(179,89,0, maxColorValue = 255))

#Order body parts as follow: Worker, Queen head, Queen thorax, Queen Abdomen, North America queen
body_part_factor <- factor(parsed_counts_sig_ratios_common_genes$body_part,
                           levels = c("W", "QH", "QT", "QA", "whole_body_queen"))
parsed_counts_sig_ratios_common_genes$body_part <- body_part_factor

#Plots
p7 <- ggplot(data = parsed_counts_sig_ratios_common_genes, aes(x = body_part, y = lfc)) + geom_boxplot(varwidth = TRUE, aes(weight = log(medians + 1.1), 
                                                                                                       fill = body_part, color = body_part), outlier.size = 0.5) +
        scale_fill_manual(values = fill_colors,
                          labels = c("Whole worker\nSouth America", "Queen Head\nSouth America", "Queen Thorax\nSouth America", "Queen Abdomen\nSouth America", "Whole queen\nNorth America"),
                          name = "Caste, body part\npopulation") +
        scale_color_manual(values = contour_colors,
                           labels = c("Whole worker\nSouth America", "Queen Head\nSouth America", "Queen Thorax\nSouth America", "Queen Abdomen\nSouth America", "Whole queen\nNorth America"),
                           name = "Caste, body part\npopulation") +
        facet_grid(. ~ gene) +
        geom_hline(yintercept = 0, color = "black") + geom_hline(yintercept = -1, color = "darkgrey", linetype = "dashed") + 
        geom_hline(yintercept = 1, color = "darkgrey", linetype = "dashed") +  
        theme_bw() + theme(panel.grid.minor = element_blank()) +
        scale_y_continuous(breaks = seq(-10, 10, 2)) +
        theme(axis.text.x = element_blank(), axis.title.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_text(vjust = 0.5, size = 12), axis.title.y = element_blank(),
              strip.text.x= element_text(size = 9), legend.title = element_text(size = 15),
              legend.text = element_text(size = 12))
        
p8 <- ggplot(data = parsed_counts_sig_ratios_common_genes, aes(x = body_part, y = lfc)) + geom_boxplot(varwidth = TRUE, aes(weight = log(medians + 1.1), 
                                                                                                       fill = body_part, color = body_part), outlier.size = 0.5) +
        scale_fill_manual(values = fill_colors,
                          labels = c("Queen Abdomen\nSouth America", "Queen Head\nSouth America", "Queen Thorax\nSouth America", "Whole worker\nSouth America", "Whole queen\nNorth America"),
                          name = "Caste, body part\npopulation") +
        scale_color_manual(values = contour_colors,
                           labels = c("Queen Abdomen\nSouth America", "Queen Head\nSouth America", "Queen Thorax\nSouth America", "Whole worker\nSouth America", "Whole queen\nNorth America"),
                           name = "Caste, body part\npopulation") +
        facet_grid(. ~ gene) +
        geom_hline(yintercept = 0, color = "black") + geom_hline(yintercept = -1, color = "darkgrey", linetype = "dashed") + 
        geom_hline(yintercept = 1, color = "darkgrey", linetype = "dashed") +  
        theme_bw() + theme(panel.grid.minor = element_blank()) +
        scale_y_continuous(breaks = seq(-10, 10, 2)) +
        theme(axis.text.x = element_blank(), axis.title.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_text(vjust = 0.5, size = 12), axis.title.y = element_blank(),
              strip.text.x= element_text(size = 9), legend.position = "none")
#Save the plot
ggsave(filename = "results/boxplot_lfcs_common_sig_only_per_gene_body_part.pdf", plot = p8, device = "pdf", height = 150, width = 297, units = "mm")
ggsave(filename = "results/boxplot_lfcs_common_sig_only_per_gene_legend_body_part.pdf", plot = p7, device = "pdf", height = 150, width = 297, units = "mm")
```

## Euler diagram for the different lists of significant genes

```{r euler_diagram}
euler_sig_genes <- euler(combinations = list("South America only" = sig_genes_sa,
                                    "North America only" = sig_genes_na_all,
                                    "Combined analysis" = sig_genes_both_lme))

#Generate plot
euler_sig_genes_plot <- plot(euler_sig_genes, quantities = TRUE, edges = TRUE, labels = list(fontsize = 8))
## Save:
ggsave(plot = euler_sig_genes_plot, filename = "results/euler_plot.pdf", width = 20, height = 20, units = "cm")
```


